---
title: "Common Marmoset Gut Microbiome Profiles in Health and Intestinal Disease"
author: "Alex Sheh and Jose Molina Mora"
date: "September 21, 2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading R data file to generate Supplemental Figures 2&3. 
Includes the following data:
*Supplemental figure 2a-d - PCA plots of microbiome data of healthy marmosets
*Supplemental figure 3a - Comparison of multiple algorithms
*Supplemental figure 3b - Evaluating stability by adding more ASVs
*Supplemental figure 3c - Box and whisker plots for top 10 ASVs
*Supplemental figure 3d - heatmap of microbiome data using top 10 ASVs selected by RF

```{r data}
load("sfig23_data.RData")
```


```{r libraries}
# for ML algorithms
library(caret)
library(ROCR)    
library(rpart)   
library(rattle)  
library(ellipse)
library(ggfortify)
library(plotrix)
library(gcrma)
library(RColorBrewer)
library(kmed)
library(DescTools)

sessionInfo()

```
# Supplemental Figure 2a-d - healthy
Uses different metadata to visualize healthy microbiome data

```{r ordination for healthy, results="hide", warning=FALSE}
#boxcox to eliminate arch effect
Data <- BoxCox(hlt_otu,0.5)
conditions <- hlt_meta
pca<-prcomp(Data)
autoplot(pca, data = conditions, colour = 'Source', main ="PCA for all data")
autoplot(pca, data = conditions, colour = 'Tissue', main ="PCA for all data")
autoplot(pca, data = conditions, colour = 'Age_at_sampling', main ="PCA for all data")
autoplot(pca, data = conditions, colour = 'Sex', main ="PCA for all data")
```
# Supplemental Figure 3
```{r ml algorithm, results="hide", warning=FALSE}

set.seed(2)
test_index<-createDataPartition(conditions$Source, p=0.80, list = FALSE)
Dtraining <- Data[test_index, ]
Dtesting<- Data[-test_index,]
Conditrain<-conditions[test_index, ]
Conditesting<-conditions[-test_index,]
  
percentage <- prop.table(table(Conditrain$Source)) * 100
percentage2 <- prop.table(table(Conditesting$Source))*100
# Run algorithms using 10-fold cross validation
control <- trainControl(method="cv", number=10, classProbs=TRUE)
metric <- "Accuracy"

#Clasification algorithms
# a) linear algorithms
#fit.lda <- train(Dtraining, Conditrain$Source, method="lda", 
#metric=metric, trControl=control)
# b) nonlinear algorithms
# CART
fit.cart <- train(Dtraining, Conditrain$Source, method="rpart", metric=metric, 
                  trControl=control)
# kNN
fit.knn <- train(Dtraining, Conditrain$Source, method="knn", metric=metric, 
                 trControl=control)
# c) advanced algorithms
# SVM
fit.svm <- train(Dtraining, Conditrain$Source, method="svmRadial", metric=metric, 
                 trControl=control)
# Random Forest
fit.rf <- train(Dtraining, Conditrain$Source, method="rf", metric=metric, 
                trControl=control)

#summarize accuracy of models
results <- resamples(list(cart=fit.cart,svm=fit.svm,knn=fit.knn,rf=fit.rf,rf=fit.rf))
# summary(results)
```
# Supplemental Figure 3a
```{r comparison algorithms}
# compare accuracy of models
dotplot(results)
```

```{r compute accuracy, results="hide", warning=FALSE}
#ALGORITMO SELECCIONADO
#model=fit.knn
#kalgoritmo="knn"
#model=fit.svm
#kalgoritmo="svmRadial"
model=fit.rf
kalgoritmo="rf"

importance <- varImp(model, scale=TRUE)
# head(importance)
# plot(importance, main = paste("All variables with algorithm", kalgoritmo))

#INDEX
IndexRank <-data.frame(sort(importance$importance$Overall, 
                            index.return = TRUE, decreasing = TRUE)[2])
Ranking<-t(IndexRank)
DtrainingRanked<-Dtraining[,Ranking[1,]]
DtestingRanked<-Dtesting[,Ranking[1,]]
DataRanked<-Data[,Ranking[1,]]

#EVALUACION DE DIF SUBSET DE variables TOP
kvalue=80

kEvaluacion<-matrix(,nrow=kvalue, ncol=52) 
colnames(kEvaluacion)<-c("Accuracy","Kappa","AccuracyLower","AccuracyUpper",
                         "AccuracyNull", "AccuracyPValue","McnemarPValue",
                         "Sensitivity_B", "Specificity_B", "Pos Pred Value_B", 
                         "Neg Pred Value_B", "Precision_B", "Recall_B","F1_B",
                         "Prevalence" ,"Detection Rate_B","Detection Prevalence_B",
                         "Balanced Accuracy_B", "Sensitivity_C", "Specificity_C", 
                         "Pos Pred Value_C", "Neg Pred Value_C", "Precision_C", 
                         "Recall_C","F1_C","Prevalence" ,"Detection Rate_C",
                         "Detection Prevalence_C","Balanced Accuracy_C",
                         "Sensitivity_E", "Specificity_E", "Pos Pred Value_E", 
                         "Neg Pred Value_E", "Precision_E", "Recall_E","F1_E",
                         "Prevalence" ,"Detection Rate_E","Detection Prevalence_E",
                         "Balanced Accuracy_E", "Sensitivity_N", "Specificity_N", 
                         "Pos Pred Value_N", "Neg Pred Value_N", "Precision_N",  
                         "Recall_N","F1_N","Prevalence" ,"Detection Rate_N",
                         "Detection Prevalence_N","Balanced Accuracy_N", "K")
k=1 
DtrainK<-as.data.frame(DtrainingRanked[,1])
colnames(DtrainK)<-colnames(DtrainingRanked)[1]
DtestK<-as.data.frame(DtestingRanked[,1])
colnames(DtestK)<-colnames(DtestingRanked)[1]

fit.algorK <- train(DtrainK, Conditrain$Source, method=kalgoritmo, metric=metric,
                    trControl=control)
predictionsK <- predict(fit.algorK, DtestK)
StatisticsK<-confusionMatrix(predictionsK, Conditesting$Source)
kEvaluacion[1,1:51]<-t(as.data.frame(c(StatisticsK$overall,StatisticsK$byClass[1,],                                    StatisticsK$byClass[2,],StatisticsK$byClass[3,],
                                       StatisticsK$byClass[4,])))

for (k in 2:kvalue){  
  DtrainK<-DtrainingRanked[,c(1:k)]
  DtestK<-DtestingRanked[,c(1:k)]
  fit.algorK <- train(DtrainK, Conditrain$Source, method=kalgoritmo, metric=metric,
                      trControl=control)
  predictionsK <- predict(fit.algorK, DtestK)
  StatisticsK<-confusionMatrix(predictionsK, Conditesting$Source)
  kEvaluacion[k,1:51]<-t(as.data.frame(c(StatisticsK$overall,StatisticsK$byClass[1,],
                        StatisticsK$byClass[2,],StatisticsK$byClass[3,],
                        StatisticsK$byClass[4,])))

}

EvalK<-as.data.frame(kEvaluacion)
```
# Supplemental Figure 3b

```{r Supplemental Figure 3b}
par(mfrow = c(1,1))
plot(EvalK$Accuracy,type="o",pch=1,col="green",
     main = paste("Evaluation by ranked variables with algorithm",kalgoritmo,sed=""),
     xlab = "Top variables",ylab = "Metrics", ylim=c(0,1))
lines(EvalK$Kappa,type = "o", pch=3, col="red")
legend("bottomright",legend=c("Accuracy","Kappa"),pch=c(1,2,3),col=c("green","red"))
```
## Boxplots of top bacteria

```{r evaluate top variables, results="hide", warning=FALSE}
ks=10
DtrainKS<-DtrainingRanked[,c(1:ks)]
DtestKS<-DtestingRanked[,c(1:ks)]

DataRankedtopK<-DataRanked[,c(1:ks)]

set.seed(3)
fit.algorKS <- train(DtrainKS, Conditrain$Source, method=kalgoritmo, metric=metric,
                     trControl=control)
importance <- varImp(fit.algorKS, scale=TRUE) 
# plot(importance, main = paste('Top ', ks, "variables with algorithm", kalgoritmo))

predictionsKS <- predict(fit.algorKS, DtestKS)
StatisticsKS<-confusionMatrix(predictionsKS, Conditesting$Source)

xR <- DtrainKS
yR <- Conditrain$Source
sxR <- DtrainingRanked[,c(6,7,8,9,10,1,2,3,4,5)]

#boxplot
par(mfrow=c(2,5))

for(i in 1:10) {
  boxplot(xR[,i]~yR, main=names(xR)[i],
          col=c("lightcoral","cyan3","darkgreen","darkorange"),ylab = " ")
}
title(paste("Top 10 variables with algorithm", kalgoritmo), outer=TRUE) 

par(mfrow=c(1,1))
```
# Supplemental Figure 3c

```{r Supplemental Figure 3c}
# box and whisker plots for each attribute
featurePlot(x=sxR, y=yR, plot="box", 
            main = paste("Top 10 variables with algorithm", kalgoritmo),
            col=c("lightcoral","cyan3","darkgreen","darkorange"))

```
# Supplemental Figure 3d

```{r heatmap}

heatmap(as.matrix(t(DataRanked[,1:ks])), scale="column", col = brewer.pal(11,"Spectral"),
        ColSideColors=paste(as.numeric(conditions$Source)+1),,labCol = FALSE)

```